officedown template

### 2024-08-02 docs/verticality_future_mkd.Rmd
methods revision

As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness. In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and result in either shifts toward greater use of lower vertical positions or local extirpation [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. Both scenarios would reduce assemblage verticality, reflecting climate-induced shifts in assemblage thermal tolerance caused by local extirpation or range shifts [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality. However, the extent to which climate change may alter global patterns of verticality is unknown.

@@

Long-term climate instability may additionally impact global patterns of verticality [@scheffersVerticalArborealityHorizontal2017]. Physiological adaptations to climate variability in the canopy may confer tolerance to extreme hot and dry conditions [@tracyNotJustSmall2010; @leahyArborealityDrivesHeat2022]. Alternatively, vertical dispersal between the ground and canopy increases opportunities for behavioral adaptations to climate fluctuations [@scheffersIncreasingArborealityAltitude2013; @scheffersVerticalArborealityHorizontal2017; @bashamVerticalStratificationCollapses2020]. However, observed impacts of historical climate instability on assemblage verticality are inconsistent [@scheffersVerticalArborealityHorizontal2017; @oliveiraVerticalStratificationInfluences2019].

As climate change causes the reorganization of species assemblages globally [@scheffersBroadFootprintClimate2016], patterns of verticality may be impacted. Functional changes to community composition are often detected in response to climate change. For example, deborealization, tropicalization, and thermophilization have occurred as community thermal tolerances increase in response to warming temperatures [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. Warming and drying climates may cause arboreal species to seek refuge at lower positions within the vertical forest structure, disperse to cooler climates toward higher latitudes or elevations, or become locally extinct. These processes will alter assemblage verticality. For example, verticality may decline in tropical lowlands where species are increasingly exposed to extreme hot and dry conditions [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. In contrast, verticality may increase toward higher latitudes or elevations where warming temperatures may release constraints on arboreality. While changes in assemblage verticality have been observed across spatial environmental gradients and in response to seasonal climatic fluctuations [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], potential impacts of climate change on verticality have not been examined at a global scale.

nbirds = nrow(birds)
nmammals = nrow(mammals)
namph = nrow(amph)
nrept = nrow(repts)

Despite widespread agreement that verticality influences patterns of biodiversity and vulnerability to climate change, global patterns of verticality and its environmental drivers have not been comprehensively described for vertebrates. Here, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with global climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes across the globe. We then examine how climate change may impact assemblage verticality.

@@

## Assemblage verticality

We mapped the verticality of vertebrate assemblages in wooded habitats using 50 km grid cells and a cylindrical equal area projection. We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al. [-@dinersteinEcoregionBasedApproachProtecting2017]. Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022]. For birds, we only used ranges that covered breeding and resident ranges, because seasonality impacts the verticality of assemblages. To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024]. This dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature. For species where imputation was necessary, we considered a species to use a given vertical stratum if the imputation probability exceeded 0.7. For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014]. We classified foraging below the water surface, around the water surface, and on the ground as 0.5, in the understory as 0.667, mid-high canopy as 0.8337, and in or above the upper canopy as 1. If species used multiple vertical habitats, vertical scores were averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that did not occur in terrestrial habitats were excluded from the analysis.

Verticality exhibited a strong positive correlation with species richness. To examine spatial patterns of verticality independently of species richness, we claculated the standard effect size (SES) of mean verticality (\$ SES = (x_i - \bar{x})/sd(x)\$, where x is a vector of expected mean verticality scores given random selection of species from a species pool) . We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.

 To minimize impacts of the positive relationship between verticality and species richness on global patterns, we calculated the standard effect size (SES) of mean assemblage verticality, which determines whether verticality is higher (positive) or lower (negative) than expected by random chance given the species richness in each cell and the vertical strategies of species in each geographic realm.

@@

We obtained present and future global climate data from CHELSA v2.1 [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Climate velocity represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011]. Canopy height was optained from [@langHighresolutionCanopyHeight2023] and vegetaion dnesiy Vegetative complexity was calculated as the product of vegetation density [@crowtherMappingTreeDensity2015] and canopy height [@langHighresolutionCanopyHeight2023]. Elevation data was obtained from *data source*. All environemtnal data was resmapled to the resolution and projection of the assemblage data.

@@

>>>>

As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness. In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. These scenarios would reduce assemblage verticality in the lowlands, reflecting climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality. While changes in assemblage verticality have been observed across spatial environmental gradients and in response to seasonal climatic fluctuations [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], potential impacts of climate change on verticality have not been examined at a global scale.

nbirds = nrow(birds)
nmammals = nrow(mammals)
namph = nrow(amph)
nrept = nrow(repts)
```

@@

@@

## Species data

We mapped the verticality and richness of vertebrate assemblages in wooded habitats using 50 km grid cells and a cylindrical equal area projection. Grid cells with fewer than five species were excluded. We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al. [-@dinersteinEcoregionBasedApproachProtecting2017]. Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022]. For birds, we used breeding plus resident ranges, because seasonality impacts the verticality of assemblages. To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024]. This dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature. For species where imputation was necessary, we considered a species to use a given vertical stratum if the imputation probability exceeded 0.7. For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014]. We classified foraging below the water surface, around the water surface, and on the ground as 0.5, in the understory as 0.667, mid-high canopy as 0.8337, and in or above the upper canopy as 1. If species used multiple vertical habitats, vertical scores were averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that did not occur in terrestrial habitats were excluded from the analysis.

Verticality exhibited a strong positive correlation with species richness. To examine spatial patterns of verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality (\$ SES = (x_i - \bar{x})/sd(x)\$, where x is a vector of expected mean verticality scores given random selection of species from a species pool). We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.

@@

We obtained present and future global climate data from CHELSA v2.1 for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density). Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality (Oliveira and Scheffers 2019). All environmental data was resampled to the resolution and projection of the assemblage data.

@@

### Variable selection
We examined relationships between mean or SES verticality and the ten environmental variables: canopy height, vegetation density, vegetation complexity, maximum temperature of the warmest month, minimum temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, precipitation seasonality, and climate velocity.  To linearize the relationships between predictor and response variables, we log-transformed climate velocity and precipitation of the driest month. We then scaled all predictor variables to mean zero and variance one. To reduce impacts of collinearity on the model, we examined VIF factors of all variables and sequentially removed one variable at a time until all VIFs were less than 5 [@wartonEcostatsDataAnalysis2022]. We chose variables to remove based on a combination of which had the highest VIF, which were most likely to directly impact physiological limits to arboreality, and which were expected to change the most in response to climate change. For example, we excluded precipitation seasonality because it was strongly correlated with dry season precipitation, has a less direct impact on physiological limits to verticality than dry season precipitation, and varies less between current and future climate conditions than dry season precipitation. This procedure resulted in a final set of seven variables that included canopy height, vegetation density, maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the wettest month, precipitation of the driest month, and climate velocity. The additive effects of these variables were included as fixed effects in the model in addition to a quadratic term for maximum temperature of the warmest month, as we expected that verticality may increase with temperature up to a certain point but then decline under extremely high temperatures.

### Model description

We used spatial generalized linear mixed-effects models (GLMMs) to evaluate the impact of environmental variables on mean and SES verticality. Spatial models were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which fits models using maximum marginal likelihood estimation through template model builder (TMB) and uses the stochastic partial differential equation (SPDE) approach to approximate spatial Gaussian random fields. [@lindgrenExplicitLinkGaussian2011] as implemented in R-INLA [@lindgrenBayesianSpatialModelling2015]. sdmTMB is also able to fit models that include a spatially varying coefficient (SVC), which allow the coefficient of a predictor variable to vary across the study extent, and are better able to detect complex species-environment relationships [@doserGuidelinesUseSpatially2024; @andersonSdmTMBPackageFast2024].

The general model can be described as:

$$
y_s ∼ 𝜇_{s}, \\
𝜇_s = 𝑓^{-1}(𝜲^{main}_s𝛽 + α_g + 𝜲^{svc}_sζ_s + ⍵_s), \\
⍵_s ∼ MVNormal(0,𝚺_⍵)
$$

where $y_s$ is mean or SES verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected mean or SES verticality, and $𝑓^{-1}$ represents an inverse link function. We modeled SES verticality as a normal distribution and mean verticality as a beta distribution with a logit link function because values are constrained between 0 and 1. $𝜲^{main}_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients. $α_g$ represents an optional random intercept by group g. $⍵_s$ represents a spatial random field at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates or random intercept. $𝜲^{svc}_sζ_s$ represents an optional spatially varying coefficient, where $ 𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$). The spatial range for the covariance function is shared with the range for the spatial random field. Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with the terms following a multivariate normal distribution with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance. The model is specified such that both spatial terms share the same range parameter.

### Mesh development

The Gaussian random field is approximated using a spatial mesh. We constructed separate meshes for all taxa and for mean and SES verticality using the ‘fmesher’ R package. The mesh used to approximate the spatial Gaussian random field (GRF) impacts model results as well as computational time. To assess sensitivity of the model to the mesh and select the optimal mesh, we used an iterative procedure to update minimum and maximum edge lengths of triangles in the mesh based on the spatial range of the data (i.e., the distance at which points exhibit a correlation of approximately 0.1) [@thorsonSpatioTemporalModelsEcologists2024]. We first estimated the spatial range by visually inspecting spatial correlations based on a subset of 1000 points. To avoid over-smoothing the GRF while or over-fitting the model and unnecessarily increasing computational demands, the maximum edge length of triangles within the modeling domain should be several times smaller than the estimated spatial range [@andersonSdmTMBPackageFast2024]. To construct the initial mesh, we specified the maximum edge length of the inner domain as 1/5 the spatial range. Additionally, we specified, an inner extension equal to the length of the spatial range, an outer extension equal to twice the length of the spatial range, and the maximum edge length in the outer extension equal to the spatial range (*cite*). To avoid small triangles, we also we specified the minimum edge length as 1/5 the maximum edge length of the inner domain [@krainskiAdvancedSpatialModeling2019]. We then fit a model using the resulting mesh. If the estimated range in the model was less than five times the maximum edge length of the spatial domain used to construct the mesh, we constructed a new mesh, updating the edge lengths and buffers based on the spatial range estimated by the model. We repeated this procedure until the spatial range estimated by the model was greater than 5 times the maximum edge length of spatial domain used to construct the mesh. We selected the mesh used in the final model for all following analysis.

### Model comparisons
